Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-11-02 manu_draft.Rmd
methods from PDV added to manu

Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as Chl-a.  However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages. STL and/or WRTDS could potentially separate seasonal from long-term trends, but doing so is not necessary to determine seasonal averages.  What is needed is an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, allowing for irregular sampling and the non-independence inherent in time-series.  This can be done with GAMs, but we develop an application that is distinct from previous studies. Even if estimates of seasonal averages and their standard errors are available, none of these methods are designed to understand interannual trends in those averages.  A Kendall test would not incorporate the standard errors or reveal useful long-term patterns beyond a significance test. Similarly, STL and WRTDS are not designed for this goal.

@@

The fundamental goal is to smooth the raw data across time in order to separate the signal from the noise. The simplest GAM for this purpose can be expressed as:

@@

Here `y` is the time-series of interest, such as Chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.

@@

This model is effectively mathematically equivalent to model S.  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.  With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.  With model SY, it is available more conveniently.  Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

@@

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.  The model will impute an average seasonal pattern where data are limited. However, a conceptual representation of what an average within-year pattern represents is challenging.  For example, suppose in every year includes a spring chl-a peak.  If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.  However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.  The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.

@@

All of the models require a sufficiently large number of knots in one or more of the spline terms to accommodate the fluctuations in the data and use the GAM estimation methods in the manner for which they were designed.  In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.  In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.  Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.  In practice, allocation of knots between the different splines can be chosen by the analyst so long it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.

@@

### Log and Box-Cox transformations

Following previous authors, we consider base 10 log-transformed chl-a levels for analysis.  To generalize this choice, we also consider Box-Cox transformations, which scale from log-transformation to no (i.e., identity) transformation as a parameter $\lambda$ scales from 0 to 1.

@@

Next we consider the goal of estimating a seasonal average, such as the mean spring Chl-A level, and its uncertainty in each year.  Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  In what follows, subscript $t$ will be omitted.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of estimate parameters and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for Chl-A is needed for a single day, given as `dec_year` = $r$, $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time; it need not be the time of an observation.

To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.  The seasonal averages used in our examples are calculated at the resolution of days.  The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.  The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$.  Since parameter estimates are correlated, it is important to use the full $\hShb$ to determine the covariance of $\hby$.  This is given by $\hShy = \bX \hShb \bX^T$.

Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of this is $\hsshm = A^T \hShy A$.  The standard error of the seasonal average is $\hshm$.  These are all standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.  @Murphy19 gave a similar summary for the purpose of comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series. [I note that equations 2-4 of Murphy et al. appear to be wrong, giving standard least-squares rather than penalized least-squares. I do not necessarily feel we should distract readers with this correction, but we could do so if desired.]

[Here are more detailed notes on @Murphy19.  I think this is more detail than necessary to include in a paper, but I note it here in case it is helpful for clarity among our group.  I'm fairly sure we are both doing the same thing, just with slightly different notation. @Murphy19 used the above relationships to set up a test of the null hypothesis that `y` is the same at two times (each potentialy an average, but not shown here).  To do so, $\bX$ would have two rows, one for each time.  The vector $A^T$ would [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in Chl-A between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.  One can then determine if the difference is significantly different from 0. For me, the most confusing statement of the explanation in @Murphy19 was to call $Z_p$ the "prediction vector".  If I follow their notation, it is the vector of spline basis function evaluations, so it is used to predict `y` but is not itself a set of predictions.]

@@

In this section we explain how to use the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, to investigate across-year trends of within-year features.  In worked examples, we use as features two seasonal averages of Chl-A, one for January-July and the other for August-December.  These periods are chosen for their relevance to California's climate. (In the *Discussion* we consider other features of interest that may be obtained from estimated GAMs.)  For any feature of interest, as long as its estimates come with associated standard errors, we give meta-analysis methods to address two kinds of questions:

@@

[A third question would be "3. Are there significant trends or non-linear fluctuations across years in a seasonal average, where the time-scale of trends is chosen by statistical criteria?"  This needs further work.  My current instinct is that this is too much for this paper, but it should be discussed.]

@@

To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.  To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test. The latter readily includes consideration of multiple testing.  For example, this is provided in R package `multcomp`.  While this package typically uses the results of a linear or generalized linear model fit as input, it is easy to set up the values of $\hmt$ and $\hshmt$ as inputs.  Details are given in the Appendix.

@@

The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19].

@@

where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$. Here the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.  The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.  This is assumed to be normally disributed with mean 0 and variance $\hsshmt$, which is assumed known from the calculations above.  The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$, what would be considered the "residual" in the sense of unexplained variation not due to the estimation error.  This is assumed normal with mean 0 and unknown variance, $\sbs$, to be estimated.

The R package *mixmeta* was used to estimate this model [@Sera19].  Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ handles this model as a linear mixed effects model with some variance components fixed.  The default estimation method is restricted maximum likelihood (REML).

We apply this method for windows of a chosen width of years.  In this way, one can ask whether there is a significant linear trend in Chl-A over a chosen series of years, taking into account all the data in those years.  Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error. The meta-analysis incorporates such differences in uncertainty magnitudes.

>>>>

Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point. Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.  However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages. STL and/or WRTDS could potentially separate seasonal from long-term trends, but doing so is not necessary to determine seasonal averages.  What is needed is an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, allowing for irregular sampling and the non-independence inherent in time-series.  This can be done with GAMs, but we develop an application that is distinct from previous studies. Even if estimates of seasonal averages and their standard errors are available, none of these methods are designed to understand interannual trends in those averages.  A Kendall test would not incorporate the standard errors or reveal useful long-term patterns beyond a significance test. Similarly, STL and WRTDS are not designed for this goal.

@@

The fundamental goal is to smooth the raw data across time to separate a signal of response variable from noise. The simplest GAM for this purpose can be expressed as:

@@

Here `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.  Following previous authors [@Murphy19;@Yang20], we consider log-transformed (base 10) chlorophyll levels for analysis.

@@

This model is effectively equivalent to model S.  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.  With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.  With model SY, it is available more conveniently.  Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

@@

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.  The model will impute an average seasonal pattern where data are limited. However, a conceptual representation of what an average within-year pattern represents is challenging.  For example, suppose in every year includes a spring chlorophyll peak.  If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.  However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.  The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.

@@

All of the models require a sufficiently large number of knots in one or more of the spline terms to accommodate the fluctuations in the data and use the GAM estimation methods in the manner for which they were designed.  In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.  In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.  Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.  In practice, allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.

@@

@@

Next we consider the goal of estimating a seasonal average, such as the mean spring chlorophyll concentrations and the estimated uncertainty in each year.  Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$. The season includes $n$ days.  For simplicity, the following text omits subscript $t$.

Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of estimate parameters and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.  Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.  For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns. The number of spline basis functions is related to the number of knots. Note that $r$ can be any time, not necessarily the time of an observation.

To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.  Here, the seasonal averages used in our examples are calculated at the resolution of days.  The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.  The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$.  Since parameter estimates are correlated, it is important to use the full $\hShb$ to determine the covariance of $\hby$.  This is given by $\hShy = \bX \hShb \bX^T$.

Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.  The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.  Each of these estimates are standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.  A similar summary for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs. end of a time-series is provided by @Murphy19.  For example, the above estimates can be used for a test of the null hypothesis that `y` is the same at two times. To do so, $\bX$ would have two rows for each time to compare.  The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.  This difference can then be used to determine if the two estimates in time are significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates.

@@

This section explains how the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, are used to investigate across-year trends of within-year features.  In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.  These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the San Francisco Bay region. For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:

@@

@@

To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.  To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test. The latter readily includes consideration of multiple testing, which is provided in the R package `multcomp` [@Hothorn08].  While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.  Details are given in the Appendix.

@@

The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19]:

@@

where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$. Here the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.  The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.  The residuls is assumed to be normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.  The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.  The random effect is assumed a normal distribution with mean 0 and unknown variance, $\sbs$, to be estimated.

The R package *mixmeta* was used to estimate the model in equation \@ref{eqn:mixmet} [@Sera19].  Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.  Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed, whereas others are estimated ($\sbs$).  The default estimation method for meta-analysis models is restricted maximum likelihood (REML).

We apply this method for windows of a chosen width of years.  This approach provide evaluates whether there is a significant linear trend in chlorophyll over a chosen series of years, by providing a more comprehensive assessment of all the data in those years (i.e., the seasonal averages, plus their uncertainty).  Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error. The meta-analysis explicitly includes these differences in uncertainty magnitudes.
